---
output:
  github_document:
    html_preview: true
---


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

<!-- badges: start -->
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/opisthokonta/implied?branch=master&svg=true)](https://ci.appveyor.com/project/opisthokonta/implied)
[![Travis build status](https://travis-ci.org/opisthokonta/implied.svg?branch=master)](https://travis-ci.org/opisthokonta/implied)
[![Codecov test coverage](https://codecov.io/gh/JonasMoss/implied/branch/master/graph/badge.svg)](https://codecov.io/gh/JonasMoss/implied?branch=master)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
<!-- badges: end -->

## implied

This package converts bookmaker odds into proper probabilities with the function 
`implied_probabilities`. Currently six methods are available, each of them 
with different assumptions about how the bookmakers convert their probabilities 
into odds.  

The natural conversion of bookmaker odds into probabilities is to take the 
inverse of the odds. This procedure has two serious problems, as the converted 
probabilities

1. The converted probabilities are not proper, as they can sum to more than one. 
   The excess probability is called the *bookmaker margin*. 
2. The converted probabilities will be biased due to 
   [favorite-longshot bias](https://en.wikipedia.org/wiki/Favourite-longshot_bias). 
   This happens even if the bookmaker margin is accounted for.
   
The methods in this package remove the bookmaker margin and some of them also 
adjust for favorite-longshot bias.

## Installation
Install the package using `devtools` from inside `R`.

```{r install, eval=FALSE}
install.packages("devtools")
devtools::install_github("opisthokonta/implied")
```


## Usage

A bookmaker provides the odds 4.20 for home, 3.70 for a draw, and 1.95 for 
away. To calculate the implied probabilities for these odds using the
`basic` method call:

```{r basic}
library(implied)

# method = c("basic", "shin", "wpo", "or", "power", "additive")
implied_probabilities(c(4.20, 3.70, 1.95), method = "basic")

```

## Documentation

See the [overview vignette](https://github.com/ophistokonta/implied/vignettes/overview)
for an overview of the package, including a description of the methods used to
calculate implied probabilities.

## How to Contribute or Get Help
If you encounter a bug, have a feature request or need some help, open a [Github issue](https://github.com/ophistokonta/implied/issues). Create a pull requests
to contribute. This project follows a [Contributor Code of Conduct](https://www.contributor-covenant.org/version/1/4/code-of-conduct.md).
