---
title: "Troubleshooting"
author: "Jonas C. Lindstrøm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In this document you can find some guidance on what to do if you get errors or weird results. 

Errors and warning messages will typically occur when the algorithms that convert 
odds to probabilities fails to give a proper results. This does not necessarily 
mean that there is a bug in the algorithms, it is just as likely that the mathematical 
relationship between the odds and the underlying probabilities does not conform to 
the assumptions needed for the different methods to work.

In my experience, the following scenarios can often cause problems for at least 
some of the methods:

- Extremely large odds can often create problems.
- Many outcomes, such win-odds for competitions with many contestants.
- Very large bookamker margins.

When the conversion from odds to probabilities fail, or give inappropriate results, 
a warning will show that indicates what the problem might be. In addition, the output 
from the implied_probabilities function will contain a vector that indicates which 
lines there are problems. 

## Identifying problematic results

Here is an example. In the code below there are 3 lines of 4-way odds that is converted
to probabilities using the 'additive' method. The first of these odds fails to be
properly converted and you get a warning saying 'Probabilities outside the 0-1 
range produced at 1 instances'.

The point of this example is not to 'fix' the results or tweak the algorithm to 
work. The algorithm works as it should, it is just that the mathematical relationship 
between the odds and the underlying probabilities does not work well with the additive
method. The point is just to show how to find the problematic results, if they occur.

The methods 'wpo', and 'bb' doesn't work with this set of odds either, but the 
rest does. 

```{r example1_1}
require(implied)

my_odds <- rbind(c(1.15, 5, 10, 25),
                c(4.1, 4.2, 8.2, 2.1),
                c(3.8, 4.7, 5.9, 2.3))

my_probs <- implied_probabilities(my_odds, method = 'additive')

```


The vector named "problematic" indicates that there is a problem in the first odds-line 
in the input. 

```{r example1_2}
my_probs$problematic

# Can also just list the line numbers
which(my_probs$problematic)

```

And if we look at the probabilities, you will see that the 4th probability 
in the first line is negative, which isn't a valid probability. 

```{r example1_3}
my_probs$probabilities
```




## Warning: Probabilities outside the 0-1 range

This warning means that some of the converted probabilities are outside the valid
range of probabilities, which is between 0 and 1. Most likely a negative probability.

Unless there is an accompanying warning about uniroot-problems, there isn't much
to do about this, and you should conclude that the conversion method you have used
is incompatible with the odds you have. Try another method. 



## Warning: Error in stats::uniroot: f() values at end points not of opposite sign

This error can happen when using the methods 'shin', 'or', 'power', or 'jsd'. These
methods convert the odds to probabilities using an equations solver called uniroot. 
Uniroot does a search of possible values of the factor used in the methods, and 
finds the factor that gives correct probabilities (ie they sum to 1). Sometimes 
the solver cant find the probabilities.

There are two possible reasons for why the solver cant find the correct factor and 
the correct probabilities. The first reason (and the most likely) is that the method 
you have chosen simply does not have a valid solution. Unfortunately, there is not 
really much to do about it, other than using a different method.

The second reason could be that some of the settings used in the solver does not 
allow the algorithm to find the solution. You can change some of the settings in
the uniroot solver via the uniroot_options argument in implied_probabilities(). 

The following uniroot settings can be changed: interval, maxit, tol and extendInt. 
Take a look at the help page for the uniroot function for more information about
the different settings. 



## Warning in log(x/y) : NaNs produced

This warning sometimes occur with method 'jsd', when the odds are extreme or otherwise
difficult to convert. This does not get flagged as problematic, because it might 
not actually be a problem. But you should check if the probabilities in question 
seem reasonable. 


## Error: Some inverse odds sum to less than 1.

This error occurs when the naive implied probabilities sum to less than 1. The whole 
point of the conversion methods in this package is to convert odds to proper probabilities
where the odds imply a total probability greater than 1, which gives the bookmaker's
an advantage. If they sum to less than one it means that the bookmaker's odds are advantageous
for the bettor. This is a very unlikely scenario, and it is most likely due to an
error in your data processing pipeline. 

The conversion methods might be made to work in this case, but I haven't tried or
tested it. This might cahnge in the future. 


## Warning: Could not find z: Did not converge in x instances. Some results may be unreliable.

This warning can happen with method 'shin', and with shin_method = 'js'. There are 
two possible fixes: 

 - Try to increase shin_maxiter from the default 1000 to something larger, like 2000. 
 - Change shin_method to 'uniroot'.


## Warning: z estimated to be negative: Some results may be unreliable. 

This warning can happen with methods 'shin' and 'bb'. I am actually not sure if
the results should be considered unreliable, or if they can be useful. These are 
not flagged as problematic, and you need to look at the 'zvalues' in the output
to see which ones are negative. 





